# When You Have Outliers or Non-Normal Distributions: Wilcoxon and Kruskal-Wallis Rank Tests


If you have outliers in your data, or the data within each treatment group are not normally distributed, t-tests are more likely to give non-significant p-values. Outliers and non-normal distributions reduce the power of t-tests to detect significant effects. Even when graphs of your data show that there is a difference among the groups, you may get non-significant p-values from t-tests as a result of the outliers or non-normal distributions.


In these cases, rank methods and robust methods are alternatives to the t-test. We’ll begin with some of the commonly used rank alternatives to t-tests and ANOVA.


We’ll see examples of each of these shortly. However, none of these methods allow you to have more than one explanatory variable. For this reason, I suggest that, instead of using these methods, you consider the methods that allow for more than one explanatory variable, described in the following chapter.


If your sample size is large, deviations from normal distribution may be less of a problem for t-tests and ANOVA. However, even with a large sample size, you can often increase power by using a robust method or rank method.


As an example of what we mean by ranks, here are observations from the sleep data frame from R with the original values of the variable extra and the values of extra converted to ranks. You can use the help(sleep) command to read a description of the data set.


```{r}
ranked.data  = data.frame(extra = sort(sleep$extra), ranked.extra = rank(sort(sleep$extra)))

ranked.data
```


```{r}
   extra ranked.extra
1   -1.6          1.0
2   -1.2          2.0
3   -0.2          3.0
4   -0.1          4.5
5   -0.1          4.5
6    0.0          6.0
7    0.1          7.0
8    0.7          8.0
9    0.8          9.5
10   0.8          9.5
11   1.1         11.0
12   1.6         12.0
13   1.9         13.0
14   2.0         14.0
15   3.4         15.5
16   3.4         15.5
17   3.7         17.0
18   4.4         18.0
19   4.6         19.0
20   5.5         20.0
```


## Alternative to two-sample t-test: Wilcoxon rank sum for two independent samples 


Let’s start with an example where we use a t-test for the question, do colon cancer patients have elevated levels of the mucin protein in their blood?  In this synthetic data, we have the level of the protein mucin in two groups: (1) patients with colon cancer and (2) healthy controls. The two-sample t-test is commonly used for such data.


Here is R code with the data and a graph. 


```{r}
cancer=c(83, 89, 90, 93, 98)
controls=c(99, 100, 103, 104,  141)
Mucin=c(cancer,controls)

Diagnosis=c(rep("Cancer",5), rep("Controls",5))
colon.data=data.frame(Diagnosis, Mucin)

colon.data
```


```{r}
   Diagnosis Mucin
1     Cancer    83
2     Cancer    89
3     Cancer    90
4     Cancer    93
5     Cancer    98
6   Controls    99
7   Controls   100
8   Controls   103
9   Controls   104
10  Controls   141
```


```{r}
with(colon.data, boxplot(Mucin ~ Diagnosis, ylab="Mucin", ylim=c(50,150)))
```


The boxplot and data show that there is complete separation between the two groups. In fact, all the cancer subjects have lower mucin values (98 or less) than any of the controls (99 or greater). But notice the large outlier in the control group, where one subject has a mucin value of 141.


Here are the means and medians of the two groups. Notice that the mean of the control group, 109.4, is larger than the median, 103. This difference is due to the outlier.


```{r}
aggregate(Mucin ~ Diagnosis, colon.data, mean)

  Diagnosis Mucin
1    Cancer  90.6
2  Controls 109.4

aggregate(Mucin ~ Diagnosis, colon.data, median)
  
Diagnosis Mucin
1    Cancer    90
2  Controls   103
```


Here is the ordinary two-sample t-test, using the R function t.test.


```{r}
with(colon.data, t.test(Mucin ~ Diagnosis, var.equal=TRUE))

        Two Sample t-test

data:  Mucin by Diagnosis
t = -2.258, df = 8, p-value = 0.05389
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -37.9994752   0.3994752
sample estimates:
  mean in group Cancer mean in group Controls 
                  90.6                  109.4
```


The boxplot and the data show that there is complete separation of the two groups, but the p-value for the t-test is not significant (p = 0.05389). The reason that the t-test is not significant is the large outlier. The outlier increases the standard deviation, which increases the standard error, which gives the non-significant p-value.


The output from the t-test includes the means of the two groups (90.6 and 109.4), and the 95% confidence interval for difference between the means of the two groups (-37.99 to 0.399). Because the 95% confidence interval includes 0, we get p > 0.05.


The Wilcoxon rank sum test is an alternative to the t-test. The Wilcoxon rank sum test uses the rank value of each observation, rather than the actual value, so it is not affected by the outlier. Here are the rank values of the original observations.


Here is the Wilcoxon test using the R function wilcox.test.


```{r}
with(colon.data, wilcox.test(Mucin ~ Diagnosis, conf.int=TRUE))

Wilcoxon rank sum test

data:  Mucin by Diagnosis
W = 0, p-value = 0.007937
alternative hypothesis: true location shift is not equal to 0
95 percent confidence interval:
 -51  -5
sample estimates:
difference in location 
                   -13
```


Using the rank values, the Wilcoxon rank sum test yields a p-value of p = 0.0079. We reject the null hypothesis and conclude that colon cancer patients have mucin levels different from those of healthy controls.


The output from the Wilcoxon test includes the sample estimate of the difference in location, which is -13 for these data. The difference in location, the Hodges-Lehmann estimator, is the median of all pairwise differences between the two groups. That is, we calculate the 5*5 = 25 pairs of differences between each value in the cancer group and each value in the control group, and then take the median of those 25 differences. 


The 95% confidence interval for the difference in location for these data is -51 to -5 (the Moses confidence interval). Because the 95% confidence interval does not include 0, we get p < 0.05. The Hodges-Lehmann estimator is not necessarily equal to the difference between the medians of the two groups. As described below, there may be no difference between the group medians. But if one group tends to have higher (or lower) values than the other group (stochastic dominance), the Wilcoxon test can give p < 0.05 even if the medians are identical.


## Sample size required for a t-test to give power of 0.8


What sample size would we need if we wanted to get a significant p-value for the t-test for the colon data? The difference between the group means is delta = 109.4-90.6 = 18.8. The standard deviation in the control group is sd = 17.8. We’ll use the usual significance level (alpha) of sig.level = 0.05, and desired power of 0.8. The function power.t.test gives the required sample size for these assumptions:


```{r}
power.t.test(delta = 18.8, sd = 17.8, sig.level = 0.05, power = 0.8)

     Two-sample t test power calculation 

              n = 15.09557
          delta = 18.8
             sd = 17.8
      sig.level = 0.05
          power = 0.8
    alternative = two.sided

NOTE: n is number in *each* group
```


We would need a sample size of 15 per group, 30 total, to have an 80% chance of getting p < 0.05 using a t-test. The chapters on sample size and power provide more explanation.


## Alternative to paired t-test: Wilcoxon signed rank test


The paired t-test is often used to test for difference in paired (matched) samples, such as the difference before and after treatment. The Wilcoxon signed rank test is a rank-based analog of the paired t-test. Here's an example.


Suppose we would like to know if bears lose weight between winter and spring, while they are hibernating. We collect the following data.


```{r}
bears.winter=c(300,470,550,760,800,955,1260,1290)
bears.spring=c(280,445,520,715,750,890,990,900)
Weight=c(bears.winter, bears.spring)
Season=c(rep("Winter",8), rep("Spring",8))
bear.data=data.frame(Season,Weight)

bear.data

Season Weight
1  Winter    300
2  Winter    470
3  Winter    550
4  Winter    760
5  Winter    800
6  Winter    955
7  Winter   1260
8  Winter   1290
9  Spring    280
10 Spring    445
11 Spring    520
12 Spring    715
13 Spring    750
14 Spring    890
15 Spring    990
16 Spring    900
```


The graph of the weight change shows that there are two large outliers.


```{r}
plot(bears.winter-bears.spring)
```


We’ll first use the paired t-test to examine the change in weight of bears, where the same bears were weighed in winter and in spring. Then we’ll analyze the data using the Wilcoxon signed rank test.


```{r}
with(bear.data, t.test(Weight ~ Season, paired=TRUE))

Paired t-test

data:  Weight by Season
t = -2.274, df = 7, p-value = 0.05714
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -228.210666    4.460666
sample estimates:
mean of the differences 
               -111.875
```


The paired t-test gives p = 0.057, indicating that bears do not lose weight while hibernating. 


```{r}
with(bear.data, wilcox.test(Weight ~ Season, paired=TRUE))

Wilcoxon signed rank test

data:  Weight by Season
V = 0, p-value = 0.007813
alternative hypothesis: true location shift is not equal to 0
```


The Wilcoxon signed rank test gives a significant result, p = 0.0078, indicating that bears lose weight during winter hibernation. 


Notice that all the bears actually lose weight. Using the paired t-test, the p-value is not significant, and we conclude that bears don’t lose weight. The Wilcoxon signed rank test gives us p =0.0078, so we reject the null hypothesis, and conclude that bears do lose weight.


## Rank alternatives to one-way ANOVA: Kruskal-Wallis and Rfit


One-way ANOVA allows us to compare two or more treatment groups. The Kruskal-Wallis test is one rank alternative. Another alternative is the rank-based oneway.rfit function from the package Rfit.


We’ll use the quail data set from the Rfit package. Thirty-nine quail were randomized to one of 4 treatments for lowering low density lipoprotein (ldl). The variable “treat” is a factor. The variable ldl is the low density lipoprotein.


```{r}
install.packages("Rfit") # if not already installed
library(Rfit)
```


```{r}
head(quail, 10) # display the first 10 rows of the quail data set

   treat ldl
1      1  52
2      1  67
3      1  54
4      1  69
5      1 116
6      1  79
7      1  68
8      1  47
9      1 120
10     1  73
```


```

```


```

```


```

```


```{r}
with(quail, boxplot(ldl ~ treat))
```


The boxplots suggest that there are differences among the 4 treatment groups. But there are large outliers that will adversely affect t-tests or ANOVA.


Here are the median values for each treatment.


```{r}
aggregate(ldl~treat, quail, median)

  treat  ldl
1     1 68.5
2     2 35.0
3     3 62.0
4     4 65.0
```


We’ll begin with a standard one-way analysis of variance.


```{r}
quail.lm.fit = with(quail, lm(ldl ~ treat))
anova(quail.lm.fit)

Analysis of Variance Table

Response: ldl
          Df Sum Sq Mean Sq F value Pr(>F)
treat      3   3189 1062.85  1.1433 0.3451
Residuals 35  32536  929.61 
```


The one-way analysis of variance indicates that there is no difference among the 4 treatments (p = 0.345). 


Next try the Kruskal-Wallis test.


```{r}
with(quail, kruskal.test(ldl, treat))
Kruskal-Wallis rank sum test

data:  ldl and treat
Kruskal-Wallis chi-squared = 7.1879, df = 3, p-value = 0.06614
```


The Kruskal-Wallis test approaches significance, with p = 0.066.


Alternatively, we can use the function oneway.rfit, from the Rfit package, to perform a rank-based analysis of the quail data. 


```{r}
quail.oneway.rfit = with(quail, oneway.rfit(ldl, treat))
quail.oneway.rfit
```


The oneway.rfit function indicates that there is a significant difference among the quail treatment groups p = 0.016394. The oneway.rfit output also shows pairwise comparisons that test for differences among the 4 treatment groups. By default, there is no p-value adjustment for multiple hypothesis testing. We can request estimates of the differences between each pair of treatments with confidence intervals adjusted for multiple comparisons by specifying method = ”tukey”.


```{r}
summary(quail.oneway.rfit, method="tukey")
Multiple Comparisons
Method Used  tukey 

  I J Estimate  St Err Lower Bound CI Upper Bound CI
1 1 2      -25 8.26704      -47.29541       -2.70459
2 1 3       -4 8.26704      -26.29541       18.29541
3 1 4       -5 8.49358      -27.90636       17.90636
4 2 3      -21 8.26704      -43.29541        1.29541
5 2 4      -20 8.49358      -42.90636        2.90636
6 3 4        1 8.49358      -21.90636       23.90636
```


For the pairwise comparisons, if the lower and upper bound CI do not include zero then the pair of treatments differ after Tukey correction for multiple hypothesis testing. Thus, in the first row, which compares treatments 1 and 2, the lower and upper confidence bounds are -47.29 to -.27, which does not include zero, so the difference is significant.


## Do the Wilcoxon tests and Kruskal-Wallis test compare the medians of groups?


If the treatment groups do not differ in shape, but differ only by a shift in their location, then the Wilcoxon tests and Kruskal-Wallis compare the medians of the treatment groups.  If the treatment groups differ in shape and location, these rank tests compare the entire distributions of the groups (that is, the ranks of all the observations, not just the median). 


You can construct two treatment groups that have identical medians, but that differ in their rank sums, and will give significant differences in the Wilcoxon or Kruskal-Wallis tests.


```{r}
Group.A=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1, 1,1,1,1,1,1,1,1)

Group.B=c(0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0,0, 2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2,2)
```


The medians of the groups are both zero but the standard deviations (shape) are different.


```{r}
median(Group.A) # 0
median(Group.B) # 0

sd(Group.A) # 0.5
sd(Group.B.B) # 1.0
```


```{r}
wilcox.test(Group.A, Group.B)

Wilcoxon rank sum test with continuity correction

data:  group.A and group.B
W = 1800, p-value = 0.01428
alternative hypothesis: true location shift is not equal to 0
```


The Wilcoxon rank sum test gives p = 0.014, indicating that the two groups are significantly different. This result shows that, although group A and group B have identical medians, they differ in scale (standard deviation), and differ significantly according to the Wilcoxon test. In this situation, the Wilcoxon test is said to be a test for “stochastic dominance”, which means that one group tends to have higher values than the other group. A caveat is that the calculation of the p-value for the Wilcoxon test assumes treatment groups do not differ in scale but differ only by a shift in their location.


## When should I use t-tests and ANOVA versus rank methods?


T-tests and the classical rank methods we looked at in this chapter may sometimes be sufficient to get good results. Their weakness is that these tests only allow for one explanatory variable. For most analyses, you are more likely to get p < 0.05 if you include additional explanatory variables in an analysis of variance, or if you use the robust and rank-based methods described in the next chapter.  However, it is still informative to compare t-tests and the rank tests. 


The Wilcoxon, Kruskal-Wallis tests don't require the assumption of normal distributions, and are not affected by outliers, so why don’t we always use them instead of t-tests and one-way ANOVA? There are several considerations.


If the data are exactly normally distributed, then ANOVA and t-tests have slightly more power to detect differences than do the standard versions of the rank tests. However, even with small or moderate deviations from normality, the rank tests will usually have power equal to or greater than the power of the tests that assume normality. When the data within treatment groups is substantially non-normal, the rank tests will pretty reliably have power equal to or greater than the tests that assume normality.


When you have a very small sample size (5 or less per group) it may be mathematically nearly impossible to get a significant p-value using a rank test. But it may be possible to get significance using a t-test or other parametric test.


One criticism of using rank methods used to be the difficulty of calculating confidence intervals. Before modern computers existed, getting confidence intervals for rank methods was harder than for methods that assume a normal distribution. But with today’s computers it is easy to get confidence intervals for rank methods.
