# When You Have Outliers and More Than One Explanatory Variable: Robust Tests

```{r}
#| include: false
library(ggplot2)
library(RobStatTM)

cancer=c(83, 89, 90, 93, 98)
controls=c(99, 100, 103, 104,  141)
Mucin=c(cancer,controls)

Diagnosis=c(rep("Cancer",5), rep("Controls",5))
colon.data=data.frame(Diagnosis, Mucin)

# install.packages("Rfit")
library(Rfit)

anova.2way.data <- tibble::tribble(
~Treatment, ~Litter, ~Response,
  "Drug", "Litter.1", 4,
  "Drug", "Litter.1", 8,
  "Drug", "Litter.2", 11,
  "Drug", "Litter.2", 12,
  "Drug", "Litter.3", 13,
  "Drug", "Litter.3", 15,
  "Drug", "Litter.4", 20,
  "Drug", "Litter.4", 21,
  "Placebo", "Litter.1", 10,
  "Placebo", "Litter.1", 15,
  "Placebo", "Litter.2", 16,
  "Placebo", "Litter.2", 17,
  "Placebo", "Litter.3", 19,
  "Placebo", "Litter.3", 20,
  "Placebo", "Litter.4", 21,
  "Placebo", "Litter.4", 23
)

anova.2way.data.outlier <- tibble::tribble(
~Treatment, ~Litter, ~Response,
  "Drug", "Litter.1", 4,
  "Drug", "Litter.1", 8,
  "Drug", "Litter.2", 11,
  "Drug", "Litter.2", 12,
  "Drug", "Litter.3", 13,
  "Drug", "Litter.3", 15,
  "Drug", "Litter.4", 20,
  "Drug", "Litter.4", 32,
  "Placebo", "Litter.1", 10,
  "Placebo", "Litter.1", 15,
  "Placebo", "Litter.2", 16,
  "Placebo", "Litter.2", 17,
  "Placebo", "Litter.3", 19,
  "Placebo", "Litter.3", 20,
  "Placebo", "Litter.4", 21,
  "Placebo", "Litter.4", 31
)
```

The Wilcoxon and Kruskal-Wallis rank tests described in the previous chapter only allow for one explanatory variable. As we saw in chapter 1, you can get much better p-values by including more than one explanatory variable. This applies to analyses where you have outliers or non-normal data.


In this chapter we’ll examine robust methods you can use for this situation. We’ll also briefly touch on rank-based methods.


Robust methods are designed to give good power when most of the observations in a data set are from a normal distribution with a particular mean and standard deviation, while a small portion of the observations are from a different distribution. The different distribution will likely have a different mean and standard deviation and might not be normal. This small portion of the observations from the different distribution are often identified as outliers.


The essential idea of robust methods is to down weight the outliers so that they have less effect on the statistical model and tests. If there are no outliers, the robust tests give results very similar to a classical t-test, analysis of variance, or linear regression. If there are outliers, the robust test will usually give better p-values than the classical tests, by giving most weight to the majority of observations.


Robust methods are available in R, JMP, SAS, Stata, S-plus and other software. The R package RobStatTM presented here accompanies the textbook “Robust Statistics Theory and Methods (with R)” by Ricardo Maronna et al. (2018). It includes recent research advances not included in other software.


## Example with one explanatory variable: Robust alternative to t-tests


The Wilcoxon test is not the only alternative when you have outliers.


Robust statistical tests down-weight outliers, so that they have less effect on the analysis and p-values. Robust statistics tests work for data with one or more explanatory variables. Here is the analysis of the colon data from Chapter 4.


Use the install.packages function to install the RobStatTM package. You only need to do this the first time you use it.


```{r}
# install.packages("RobStatTM")
```


Each time you start an R session, you need to load the library, using the command library(RobStatTM).


We’ll use the colon data that we used in the previous chapter.


```{.r-output}
colon.data
```


Perform the analysis using the function lmrobM.


```{r}
summary(with(colon.data, lmrobM(Mucin ~ Diagnosis)))
```


```{.r-output}
Coefficients:
                  Estimate Std. Error t value Pr(>|t|)    
(Intercept)         90.607      1.999  45.336 6.19e-11 ***
DiagnosisControls   10.893      2.993   3.639   0.0066 ** 
```


The  lmrobM robust test gives a p-value of 0.0066, similar to the p-value for the Wilcoxon rank sum test (p = 0.0079) that we saw in the previous chapter.


## Example data set with two explanatory variables


Now let’s look at an example that illustrates robust analysis of a data set with two explanatory variables. I’ll start with a dataset that has no outliers. I’ll then modify the data set to have outliers. I’ll show the analysis with and without outliers for an ordinary ANOVA and for the robust analysis.


We have the following (synthetic) data for an experiment with mice, where treatment was either drug or placebo, and there were 4 mice from each litter. Within each litter, 2 mice were randomly assigned to drug and 2 were randomly assigned to placebo.


```{r}
anova.2way.data
```


```{.r-output}
   Treatment   Litter Response
1       Drug Litter.1        4
2       Drug Litter.1        8
3       Drug Litter.2       11
4       Drug Litter.2       12
5       Drug Litter.3       13
6       Drug Litter.3       15
7       Drug Litter.4       20
8       Drug Litter.4       21
9    Placebo Litter.1       10
10   Placebo Litter.1       15
11   Placebo Litter.2       16
12   Placebo Litter.2       17
13   Placebo Litter.3       19
14   Placebo Litter.3       20
15   Placebo Litter.4       21
16   Placebo Litter.4       23
```


Here is a graph of the response by treatment group.


```{r}
ggplot(anova.2way.data, 
aes(y=Response, x=Treatment)) + 
geom_point() + ggtitle("Response by treatment group")
```


Here is the t-test for response versus treatment group.


```{r}
t.test(Response ~ Treatment, var.equal=TRUE, data = anova.2way.data)
```


```{.r-output}
Two Sample t-test

data:  Response by Treatment
t = -1.8664, df = 14, p-value = 0.08307
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -9.9398424  0.6898424
sample estimates:
   mean in group Drug mean in group Placebo 
               13.000                17.625 
```


The t-test is not significant, with p = 0.083. We would conclude that the drug has no significant effect. The t-test analysis can also be done using the R function lm(), which performs a linear regression.


```{r}
summary(lm (Response ~ Treatment, data = anova.2way.data))
```


```{.r-output}
Call:

lm(formula = Response ~ Treatment, data = anova.2way.data)
```


```{.r-output}
Residuals:
    Min      1Q  Median      3Q     Max 
-9.0000 -2.1562 -0.3125  2.6250  8.0000 

Coefficients:
                 Estimate Std. Error t value Pr(>|t|)    
(Intercept)        13.000      1.752   7.419 3.26e-06 ***
TreatmentPlacebo    4.625      2.478   1.866   0.0831 .  
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 4.956 on 14 degrees of freedom
Multiple R-squared:  0.1992,Adjusted R-squared: 
```


The linear regression gives the same result as the t-test (as it should) with p = 0.083.


However, the researcher thought that mice from different litters might have different responses. Mice from a litter with few kits tend to be larger, while mice from a litter with many kits tend to be smaller. This might affect their response to the drug. Here is a plot of response versus treatment, broken out by litter.


```{r}
ggplot(anova.2way.data, aes(y=Response, x=Treatment)) + geom_point() +
  facet_wrap(~Litter) +
        ggtitle("Response to treatment  \nbroken out by Litter") + ylim(0,25)
```


The graph suggests that there is some difference among the 4 litters. Litter 1 has low values for both drug and placebo, while litter 4 has high values for both. 


Run a two-way ANOVA including both treatment and litter as explanatory variables.


```{r}
lm.2way = lm(Response ~ Litter + Treatment, data= anova.2way.data)
anova(lm.2way)
```


```{.r-output}
Response: Response
          Df  Sum Sq Mean Sq F value    Pr(>F)    
Litter     3 303.187 101.062  27.323 2.135e-05 ***
Treatment  1  85.562  85.562  23.132 0.0005449 ***
Residuals 11  40.688   3.699 
```


The two-way ANOVA shows that both litter (p = 0.000021) and treatment (p = 0.00054) have a significant effect on the response. Analyses that included only treatment as the explanatory variable would lead to the wrong conclusions. What if, in addition, there were outliers in the data?


Here is the modified data set. I’ve changed the highest values in litter 4 to be 32 for the mouse receiving drug and 31 for the mouse receiving placebo. These are rows 8 and 16 in the table below.


```{r}
anova.2way.data.outlier
```


```{.r-output}
   Treatment   Litter Response
1       Drug Litter.1        4
2       Drug Litter.1        8
3       Drug Litter.2       11
4       Drug Litter.2       12
5       Drug Litter.3       13
6       Drug Litter.3       15
7       Drug Litter.4       20
8       Drug Litter.4       32
9    Placebo Litter.1       10
10   Placebo Litter.1       15
11   Placebo Litter.2       16
12   Placebo Litter.2       17
13   Placebo Litter.3       19
14   Placebo Litter.3       20
15   Placebo Litter.4       21
16   Placebo Litter.4       31
```


Here is the graph of response by treatment group for the data with outliers.


```{r}
ggplot(anova.2way.data.outlier, aes(y=Response, x=Treatment)) + geom_point() + 
ggtitle("Response by treatment group \nwith outliers")
```


Here is the t-test for the data with the outliers, run first using the t.test function and then using the lm function.


```{r}
t.test(Response ~ Treatment, var.equal=TRUE, data = anova.2way.data.outlier)
```


```{.r-output}
Two Sample t-test

data:  Response by Treatment
t = -1.1478, df = 14, p-value = 0.2703
alternative hypothesis: true difference in means is not equal to 0
95 percent confidence interval:
 -12.191454   3.691454
sample estimates:
   mean in group Drug mean in group Placebo 
               14.375                18.625
```


```{r}
summary(lm (Response ~ Treatment, data = anova.2way.data.outlier))
```


```{.r-output}
Call:

lm(formula = Response ~ Treatment, data = anova.2way.data.outlier)
```


```{.r-output}
Residuals:
    Min      1Q  Median      3Q     Max 
-10.375  -3.438  -1.500   1.625  17.625 

Coefficients:
                 Estimate Std. Error t value Pr(>|t|)    
(Intercept)        14.375      2.618   5.490 7.96e-05 ***
TreatmentPlacebo    4.250      3.703   1.148     0.27    
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Residual standard error: 7.405 on 14 degrees of freedom
Multiple R-squared:  0.08601,Adjusted R-squared:  0.02073 
F-statistic: 1.317 on 1 and 14 DF,  p-value: 0.2703
```


The t-test and linear model both show that the drug effect is not significant, p = 0.27. Let’s look at the results broken out by litter.


```{r}
ggplot(anova.2way.data.outlier, aes(y=Response, x=Treatment)) + geom_point() +
  facet_wrap(~Litter) +
ggtitle("Response to treatment  \nbroken out by Litter, with outliers")
```


The outliers in litter 4 are evident, as is the large difference in response between the litters.


We’ll first try an ordinary two-way ANOVA for this data with outliers.


```{r}
lm.2way.outlier = lm(Response ~ Litter + Treatment,
  data= anova.2way.data.outlier)
anova(lm.2way.outlier)
```


```{.r-output}
Analysis of Variance Table

Response: Response
          Df Sum Sq Mean Sq F value    Pr(>F)    
Litter     3 596.50 198.833 12.7718 0.0006627 ***
Treatment  1  72.25  72.250  4.6409 0.0542426 .  
Residuals 11 171.25  15.568   
```


The ordinary two-way ANOVA confirms the additional variability due to litter (p = 0.00066) but indicates that the treatment is not significant (p = 0.054). 


If the treatment has an effect in one litter, but a (significantly) different effect in another litter, there is an interaction. An interaction can occur if, for example, the effect of treatment differs significantly by sex, by age, or by other variables. Examples of interactions are in the Chapter ”Avoid one-variable-at-a-time experiments. Use factorial experiment design”.


Here is the test for an interaction between treatment and litter. The formula “Response ~ Litter * Treatment” specifies the interaction.


```{r}
anova(lm(Response ~ Litter * Treatment, data= anova.2way.data.outlier))
```


```{.r-output}
Analysis of Variance Table

Response: Response
                 Df Sum Sq Mean Sq F value   Pr(>F)   
Litter            3 596.50 198.833 10.8950 0.003377 **
Treatment         1  72.25  72.250  3.9589 0.081811 . 
Litter:Treatment  3  25.25   8.417  0.4612 0.717073   
Residuals         8 146.00  18.250 
```


The interaction test is not significant (p = 0.717), indicating that there is no interaction between treatment and litter. So, we can drop the interaction and use the results for the test of litter + treatment shown above.


## Robust regression using the RobStatTM package


Now we’ll do a robust analysis of the same data using the RobStatTM package. 


To run these examples, install and load the package with this code:


Here is the analysis of the outlier data, with litter and treatment as the two explanatory variables. The function lmrobM() performs the robust regression analysis. (The two-way ANOVA can be implemented as a linear regression.)


```{r}
anova.2way.data.outlier.lmrobM = lmrobM(Response ~ Litter + Treatment,
  data = anova.2way.data.outlier)

summary(anova.2way.data.outlier.lmrobM)
```


```{.r-output}
Call:

lmrobM(formula = Response ~ Litter + Treatment, data = anova.2way.data.outlier)
```


```{.r-output}
Residuals:
    Min      1Q  Median      3Q     Max 
-2.7439 -0.6934  0.5000  1.4406 13.9942 

Coefficients:
                 Estimate Std. Error t value Pr(>|t|)    
(Intercept)         6.744      1.124   6.000 8.92e-05 ***
LitterLitter.2      4.762      1.403   3.393 0.006000 ** 
LitterLitter.3      7.512      1.403   5.354 0.000233 ***
LitterLitter.4     11.262      1.714   6.569 4.03e-05 ***
TreatmentPlacebo    4.988      1.060   4.704 0.000646 ***
---
Signif. codes:  0 ‘***’ 0.001 ‘**’ 0.01 ‘*’ 0.05 ‘.’ 0.1 ‘ ’ 1

Robust residual standard error: 2.986 
Multiple R-squared:  0.6808,Adjusted R-squared:  0.5647 
Convergence in 8 IRWLS iterations
```


The robust regression indicates that treatment has a significant effect (p = 0.0006). Compare this result to the conventional two-way ANOVA result of p = 0.054 for the outlier data, which indicated that treatment was not significant.


We can also examine the results for the robust analysis for the data with no outliers.


```{r}
anova.2way.data.lmrobM = lmrobM(Response ~ Litter + Treatment,
  data = anova.2way.data) # control=cont)
summary(anova.2way.data.lmrobM)
```


```{.r-output}
Coefficients:
                 Estimate Std. Error t value Pr(>|t|)    
(Intercept)        6.9258     1.0767   6.433 4.86e-05 ***
LitterLitter.2     4.7678     1.3628   3.498 0.004984 ** 
LitterLitter.3     7.5186     1.3727   5.477 0.000193 ***
LitterLitter.4    12.0215     1.3599   8.840 2.50e-06 ***
TreatmentPlacebo   4.6127     0.9664   4.773 0.000578 ***
```


We see that, for the data with no outliers, the p-value for the robust method is similar to the p-value for the classical analysis of variance (p = 0.000545). These results indicate that, for data with no outliers, the robust method gives results very similar to the results for a classical analysis of variance. But when there are outliers, the robust method will usually give much better p-values.


## Rank-based analysis using the Rfit package


Rank-based methods convert the data to rank-based scores. We’ll do a rank-based analysis of the same data using the Rfit package. This package accompanies the textbook “Nonparametric Statistical Methods Using R” by John Kloke and Joseph W. McKean (2015) Chapman & Hall/CRC.


To run these examples, install and load the package with this code:


Here is the analysis of the outlier data, with litter and treatment as the two explanatory variables. The function rfit() performs the analysis.


```{r}
anova.2way.data.outlier.rfit = with(anova.2way.data.outlier,
  rfit(Response ~ Litter + Treatment))

summary(anova.2way.data.outlier.rfit)
```


```{.r-output}
Coefficients:
                 Estimate Std. Error t.value   p.value    
(Intercept)        7.2500     2.0287  3.5737  0.004366 ** 
LitterLitter.2     4.2500     2.5213  1.6857  0.119987    
LitterLitter.3     7.2500     2.5213  2.8755  0.015092 *  
LitterLitter.4    16.0000     2.5213  6.3460 5.475e-05 ***
TreatmentPlacebo   5.0000     1.7828  2.8046  0.017135 *  

Multiple R-squared (Robust): 0.7677992 
Reduction in Dispersion Test: 9.0932 p-value: 0.0017 
```


The rank-based test indicates that treatment has a significant effect (p = 0.017). Again, compare this result to the conventional two-way ANOVA result of p = 0.054, which indicated that treatment was not significant.


The output also indicates that there are significant differences among the litters. 


We could have started the analysis with an interaction test, using the Rfit function raov. As for the ANOVA, the formula “Response ~ Litter * Treatment” specifies the interaction test.


```{r}
with(anova.2way.data.outlier, raov(Response ~ Litter * Treatment))
```


```{.r-output}
Robust anova Table
                 DF       RD  Mean RD       F p-value
Litter            3 55.90170 18.63390 6.69423 0.01423
Treatment         1  7.37902  7.37902 2.65092 0.14214
Litter:Treatment  3  1.11803  0.37268 0.13388 0.93711
```


The interaction test is not significant (p = 0.937), indicating that there is no interaction between treatment and litter. So, the model without interaction was suitable.


We can also examine the results for rfit for the data with no outliers.


```{r}
summary(rfit(Response ~ Litter + Treatment, data = anova.2way.data))
```


```{.r-output}
Coefficients:
                 Estimate Std. Error t.value   p.value    
(Intercept)        7.2500     1.4574  4.9746  0.000419 ***
LitterLitter.2     4.5000     1.7165  2.6216  0.023761 *  
LitterLitter.3     7.5000     1.7165  4.3693  0.001119 ** 
LitterLitter.4    12.0000     1.7165  6.9909 2.297e-05 ***
TreatmentPlacebo   4.5000     1.2138  3.7075  0.003456 **
```


We see that, for the data with no outliers, the p-value for the rfit method (p = 0.003456) is similar to the p-value for the classical analysis of variance (p = 0.000545), although not quite so good. This illustrates the trade-off when using rank-based methods. The p-values for the rank-based method may not be as good as the p-values for a conventional ANOVA when the data are approximately normally distributed with no outliers. But when the distribution deviates greatly from that, the rank-based method can give better results.


## Which method should you use?


There is no single method that will always work best for all data. The robust and rank-based methods presented in this chapter will work well in many, perhaps most, situations. 


The robust methods are designed to work well when most of the data are from a normal distribution and a small portion of the data is outliers, possibly from a normal distribution with a different mean. In these situations, I find that the robust methods have very good power and usually give the best p-values.


The rank-based methods are designed to work well regardless of the distribution of the data. So, when the data have extremely unusual distributions, the rank-based methods may give better power.


The answer also depends, in part, on the software you have available and your experience with the methods. 


For detailed comparisons of the strengths and weaknesses of many rank and robust methods see the textbooks cited in the Additional Resources section at the end of this chapter, in particular the text by Rand Wilcox. However, be forewarned that these texts require a relatively advanced knowledge of statistics.


The robust methods and rank-based methods have the advantages of available software and textbooks and perform well in many data sets. However, many other methods are available. This is an active area of research in statistics. 


## Additional Resources


Ricardo A. Maronna, R. Douglas Martin, Victor J. Yohai, Matías Salibián-Barrera (2018). Robust Statistics Theory and Methods (with R), 2nd Edition. This book describes the RobustTM package.


John Kloke and Joseph W. McKean (2015). Nonparametric Statistical Methods Using R. Chapman & Hall/CRC. This book describes the Rfit package. 


Rand R. Wilcox (2022). Introduction to Robust Estimation and Hypothesis Testing (Statistical Modeling and Decision Science) 5th Edition. Academic Press.



