# Avoid Chi-Squared Tests When You Have Ordered Categories: Use CMH Tests Instead

```{r}
#| include: false
# install.packages(“vcdExtra”)
library(vcdExtra)

ordered.columns.table = array(
  data = c(1, 5, 3, 3, 5, 1),
  dim = c(2, 3),
  dimnames = list(
    Treatment = c("drug", "placebo"),
    Result = c("no remission", "partial remission", "complete remission")
  )
)

ordered.rows.and.columns.table = array(
  data = c(5, 2, 1, 3, 3, 3, 2, 4, 6),
  dim = c(3, 3),
  dimnames = list(
    Treatment = c("placebo", "low dose", "high dose"),
    Result = c("no remission", "partial remission", "complete remission")
  )
)
```

When you have tables with ordered categorical variables, the Cochran-Mantel-Haenszel (CMH) test for ordered categories will usually give smaller p-values than a chi-squared or Fisher Exact test.


Here are two examples of tables with ordered categorical variables, showing counts of the number of subjects in each category.

- A clinical trial of a cancer drug in which the possible outcomes are the ordered categories "no remission", "partial remission", and "complete remission".

| | No remission | Partial remission | Complete remission |
|---|---|---|---|
| Drug | 1 | 5 | 3 |
| Placebo | 3 | 5 | 1 |


- A clinical trial of a cancer drug in which the rows are the ordered treatment categories "placebo", "low dose", "high dose", and the columns are the ordered categories "no remission", "partial remission", and "complete remission". Both rows and columns are ordered categories.

| | No remission | Partial remission | Complete remission |
|---|---|---|---|
| Placebo | 5 | 3 | 2 |
| Low dose | 2 | 3 | 4 |
| High dose | 1 | 3 | 6 |



Published articles sometimes report non-significant results when the researchers use a chi-squared tests for tables that have ordered categories. The CMH test would provide more power and better p-values for these analyses.


The Cochran-Mantel-Haenszel (CMH) test extends the chi-squared and Fisher Exact tests in two ways:

- tables with more than 2 factors (stratified analyses)
- tables with ordered rows, columns, or both

Cochran-Mantel-Haenszel (CMH) test


We'll use the CMHtest() function from vcdExtra package.

```{r}
library(vcdExtra)
```



## Example: CMH test for table with ordered columns


Consider a clinical trial of a cancer drug versus a placebo in which the possible outcomes are the ordered categories "no remission", "partial remission", and "complete remission".


Here is the data set.

```{r}
ordered.columns.table = array(
  data = c(1, 5, 3, 3, 5, 1),
  dim = c(2, 3),
  dimnames = list(
    Treatment = c("drug", "placebo"),
    Result = c("no remission", "partial remission", "complete remission")
  )
)
```



Here is the resulting table.


```{r}
ordered.columns.table
```


```{.r-output}
         Result
Treatment no remission partial remission complete remission
  drug               1                 3                  5
  placebo            5                 3                  1
```


The chi-squared test and Fisher Exact test ignore the ordering of the columns and return non-significant p-values for this example. Let’s start with the chi-squared test, which is commonly used.


```{r}
chisq.test(ordered.columns.table)
```


The sample size is quite small, so the Fisher test is preferred over the chi-squared test. The chi-squared output gives a warning that the chi-squared approximation may be incorrect. Next try the Fisher Exact test.


```{r}
fisher.test(ordered.columns.table)
```


We get a non-significant result for both the Fisher Exact test (p = 0.1135) and the chi-squared test (p = 0.069), but this is not reliable due to small sample size.


In contrast, the Cochran-Mantel-Haenszel (CMH) test uses the fact that there are ordered rows, columns, or both, which increases the power. In this example, it gives a significant result.


```{r}
CMHtest(ordered.columns.table)
```


Here is the interpretation of the output.

- The row labeled "cor" is used for tables where both rows and columns are ordered. We’ll look at an example shortly.
- The row labeled "rmeans" (highlighted in yellow) shows the result testing that the row mean scores differ for drug versus placebo. The CMH test gives p = 0.0248, indicating that drug and placebo differ significantly in outcomes.
- The row labeled "cmeans" would apply to tables in which the rows are ordered but columns are not ordered.
- The row labeled "general" is similar to the chi-squared and Fisher tests.

The p-value of 0.0248 for the test that row means differ indicates that there is a significant difference between the drug and placebo groups.


## Example: CMH test for table with ordered rows and columns


Consider a table showing results of a clinical trial of a cancer drug as follows.

- Rows are the ordered treatment categories "placebo", "low dose", and "high dose".
- Columns are the ordered categories "no remission", "partial remission", and "complete remission".
- Both rows and columns are ordered categories.

